Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-05-20 text/manuscript-full-KS.qmd
working through results. again

-->
```
# Findings

@@

@fig-chicago-maps presents the seven Geomarkets in the Chicago metropolitan area by total population across the three census periods. The maps in the figure show how Chicago Geomarkets reflect the classic urban geographical structure of an inner-city core surrounded by a ring of suburbs as described by @RN4982 [p. 12]. Based on the color gradient scale representing total population density, @fig-chicago-maps indicates northern suburb Geomarkets in the metropolitan area - Chain of Lakes, Northwest Suburbs, Northshore, Evanston and Skokie - experienced relatively little population growth from 1980 through 2020. In contrast, the most significant population increases occured in the Western Suburbs (1.2 Million to 1.6 Million) and the South and Southwest Suburbs (less than 1 Million to 1.3 Million) Geomarkets. The City of Chicago Geomarket, forming the metropolitan area's central core, was the only locality to experience a population decline, from 3.1 Million in 1980 to 2.8 Million in 2020, despite remaining the most densely populated Geomarket.

These modest population shifts lead to substantial change in the racial/ethnic composition within and across Geomarkets over time. @fig-chicago-rq1-race illustrates the racial and ethnic composition of each Geomarket across the three census periods. In 1980, nearly all Geomarkets were predominantly White, with proportions ranging from 83% in the South and Southwest Suburbs to approximately 93% in North Shore. In contrast, the City of Chicago had a substantially more diverse population, with White residents comprising only 46% of its total population. By 2020, several Geomarkets that were previously predominantly White shifted to a more heterogeneous racial/ethnic distribution, including Chain of Lakes (59% White, 8% Black, 24% Hispanic; 7% Asian Pacific Islander), Northwest Suburbs (62% White, 3% Black, 18% Hispanic; 15% Asian Pacific Islander), Evanston and Skokie (64% White, 7% Black, 10% Hispanic; 16% Asian Pacific Islander), Western Suburbs (56% White, 9% Black, 23% Hispanic; 9% Asian Pacific Islander), and South and Southwest Suburbs (53% White, 24% Black, 18% Hispanic; 3% Asian Pacific Islander). For many Geomarkets, this balance in racial/ethnic composition was the result of large increases in the Latinx population from 1980 to 2020, which ranged from a 7 percentage point increase for the Evanston and Skokie Geomarket to a 19 percentage point increase for Western Suburbs. Some Geomarkets also experienced a substantial increase in Black populations (South and Southwest Suburbs) and in Asian populations (Northwest Suburbs and Evanston and Skokie). @fig-chicago-rq1-race also indicates that, although the North Shore and City of Chicago Geomarkets underwent some changes in racial and ethnic composition, their overall demographic patterns remained relatively the same across the three census periods. <!-- For instance, North Shore experienced some increases in Hispanic (2% in 1980 to 8% in 2020), Asian Pacific Islander (7% in 2000 to 10% in 2020), and Black (2% in 1980 to 3% in 2020) populations yet remained a predominantly White Geomarket (74% in 2020). City of Chicago also experienced relatively modest declines in its Black population (37% in 1980 to 33% in 2020), alongside increases in its Hispanic (15% in 1980 to 26% in 2020) and Asian Pacific Islander populations (4% in 2000 to 6% in 2020), further substantiating the Geomarket as the most racially/ethnic diverse in the metropolitan area. -->

We also explore socioeconomic demographics - median income, percent living below poverty, and educational attainment - within and across Geomarkets over time in @fig-chicago-rq1-ses. Overall, median income both within and across Geomarkets in the Chicago metropolitan area remained relatively stable across the three census periods. For instance, in 1980, median household income across the Chicago metropolitan area ranged from a high of \$158,000 in the North Shore Geomarket to a low of \$66,000 in the City of Chicago Geomarket. However, median incomes in all other Geomarkets were closer to the upper thresholds of affluence, such as \$120,000 in the Northwest Suburbs, \$119,000 in Evanston and Skokie, \$108,000 in Chain of Lakes, \$105,000 in the Western Suburbs, and \$99,000 in the South and Southwest Suburbs. While overall income patterns across Geomarkets remained consistent through 2020, a median income increase for the City of Chicago Geomarket (\$79,000) and decrease in the South and Southwest Suburbs Geomarket (\$93,000) narrowed the income gap between the lowest and second-lowest Geomarkets to just \$14,000.

@fig-chicago-rq1-ses also shows that from 1980 to 2020, the percentage of residents living in poverty declined in the City of Chicago Geomarket while it increased in all other Geomarkets, resulting in a narrowing of poverty gaps across the Chicago metropolitan area. Similar patterns are evident in educational attainment across Geomarkets. In 1980, the City of Chicago had the lowest percentage of residents with a bachelor's degree or higher (13%). However, by 2020, the Geomarket experienced the largest percentage point increase in college degree attainment, reaching 38%. In comparison, all other Geomarkets experienced smaller gains in the share of residents with a bachelor’s degree or higher over the same period. These trends contributed to a narrowing of the educational attainment gap across Geomarkets from 1980 to 2020.

Demographic shifts overtime have transformed the Chicago metropolitan area's Geomarket landscape from largely a binary structure into a more tiered dynamic shaped by substantial racial/ethnic and socioeconomic inequality between Geomarkets. The original Geomarket boundaries delineated significant racial and income segregation in the 1980s. This included Black, Hispanic, and Asian residents primarily in the City of Chicago Geomarket and White residents in Geomarkets forming the suburban ring of the metropolitan area. Income, poverty, and educational attainment also created a static socioeconomic divide in 1980, with southern Geormarkets (City of Chicago and South and Southwest Suburbs) characterized by lower affluence and educational attainment compared to the higher socioeconomic standing of the northwestern suburbs. However, declining White populations, Hispanic and Asian population growth, and Black out-migration, as well as narrowing of education and poverty gaps within some Geomarkets, have contributed to a more stratified Geomarket dynamic in the metropolitan area with significant disparities remaining between Geomarkets. The City of Chicago Geomarket remains the critical geographical hub across all Geomarkets for racial/ethnic diversity; however, other Geomarkets now also include substantive numbers of residents across racial/ethnic groups (such as South and Southwest Geomarket for Black residents, Chain of Lakes for Hispanic residents, and Northwest Suburbs for Asian residents). While some stark socioeconomic gaps within Geomarkets have narrowed since 1980, significant geographical income segregation remains between Geomarkets despite a more clearly defined middle-to-high socioeconomic Geomarket group (Chain of Lakes, Northwest Suburbs, Western Suburbs).

@@

To examine how the socioeconomic and racial composition of included versus excluded prospects varies when student list purchases filter on particular Geomarkets, we analyze six orders placed by a research university targeting prospects across the entire state of Illinois. The first two orders filter on the same SAT thresholds (ranging from 1020 to 1150), the same GPA thresholds (ranging from B- to A+), and filter on the high school graduating classes of 2020 and 2021 graduating classes, respectively. The remaining four orders used similar GPA and high school class filters but differed in SAT thresholds. Two orders targeted students with relatively middle-range scores (1160 to 1300) and two orders targeted those with relatively high scores (1310 to 1600). Because all six orders included an Illinois state filter, we can utilize prospects from the Chicago metropolitan area within the resulting student lists to examine Geomarket contributions by simulating the application of a Geomarket filter across low-, middle-, and high-range SAT threshold orders.

@@

The City of Chicago Geomarket increasingly becomes the only Geomarket with substantial shares of Black, and Hispanic prospects as PSAT scores increase. All Geomarkets experience growth in the shares of White and Asian prospects, while Black and Hispanic prospect shares decline. For instance, the City of Chicago Geomarket pool is 18% Asian, 7% Black, 21% Hispanic, 5% Multiracial, making it the only Geomarket that is not predominantly White (49%). Prospect pools from all other Geomarkets in this high PSAT order are predominantly White and Asian, with Black and Hispanic shares ranging from 0.4% (North Shore) to 5% (South and Southwest Suburbs) of prospects and 6% (Northwest Suburbs) to 10% (South and Southwest Suburbs) of prospects, respectively. <!--# [LEDE -- BUT IN BOTH THE MEDIUM AND HIGH THRESHOULD, CHICAGO IS THE ONLY GEOMARKET WHERE A SUBSTANTIAL SHARE OF PROSPECTS ARE NOT WHITE OR ASIAN] -->

We also explore contributions across Geomarkets to overall prospect pools by race/ethnicity. For instance, @fig-chicago-rq2-midSAT-race shows the contributions of each Geomarket to the pool of prospects from the two orders targeting middle range SAT scores (1160-1300) by race/ethnicity. The top panel presents the full distribution of the 13,643 total prospects (whose race/ethnicity is known) included in these orders by Geomarkets: 11% reside in Chain of Lakes, 13% in Northwest Suburbs, 5% in North Shore, 5% in Evanston and Skokie, 17% in City of Chicago, 28% in Western Suburbs, and 20% in South and Southwest Suburbs. These overall distributions are compared to those in the lower panels of @fig-chicago-rq2-midSAT-race, which illustrate the proportional representation of each race/ethnicity group across Geomarkets.

@@

@fig-chicago-rq2-highSAT-combo shows the magnitude of Geomarket contribution disparities increases at the highest SAT score range (1310-1600), partly driven by the relatively fewer number of first-generation college students included within these orders. Again, the City of Chicago Geomarket disproportionately contributes a disproportionate share of first-generation Asian college students (43% across both parental education categories) and first-generation Hispanic college students (52% across both parental education categories) in comparison to the overall distribution of first-generation college students within each racial/ethnic category (13% for Asian and 40% for Hispanic, respectively). Conversely, nearly all other Geomarkets contribute a disproportionately larger share of non-first generation college students across all racial/ethnic categories. -->
```
For instance, @fig-chicago-rq2-midSAT-race shows the nearly 1,000 Black students whose contact information was purchased in these middle SAT score orders were largely contributed to the overall pool by the City of Chicago and South and Southwest Suburbs. These two Geomarkets contributed disproportionately higher shares, 48% and 30%, respectively, compared to their overall contributions of 17% and 20% of all included prospects. On the other hand, all other Geomarkets contributed substantially smaller shares of Black prospects. The City of Chicago Geomarket also contributed larger shares of the nearly 3,000 Hispanic students (39% versus 17% for all prospects), while all other Geomarkets contributed comparable or relatively smaller proportions of Hispanic prospects than their overall representation among all included prospects.

In contrast, these representational patterns by Geomarket are reversed for White and Asian students. @fig-chicago-rq2-midSAT-race shows nearly all Geomarkets contributed to the pool of White prospects at rates proportional to their overall contributions to all included prospects (except City of Chicago). Similar patterns are observed for Asian students, with proportional contributions by Geomarket largely consistent with their overall distribution, except for Northwest Suburbs (19% versus 13% for all prospects) and Evanston and Skokie (10% versus 5% for all prospects) contributing disproportionately larger shares and the South and Southwest Suburbs (8% versus 10% for all prospectq8s) contributing smaller shares relative to their overall contribution to the larger prospect pool.

@@

Relative to @fig-chicago-rq2-midSAT-ses, appendix figures \@fig-chicago-rq2-midSAT-ses and @fig-chicago-rq2-highSAT-ses demonstrate Geomarket contribution patterns by first-generation college student status are less pronounced in low-SAT range orders (1020-1150) and more pronounced in high-range SAT orders (1310-1600), respectively.

Finally, we examine Geomarket contributions to prospect pools by both first-generation college student status and race/ethnicity in the Chicago metropolitan area. @fig-chicago-rq2-midSAT-combo presents the Geomarket distribution of the 12,981 total prospects (whose first-generation status and race/ethnicity is known) included in middle-range SAT (1160-1300) orders. @fig-chicago-rq2-midSAT-combo shows the City of Chicago Geomarket consistently contributes a disproportionately larger share of first-generation college students whose parents did not attend college across all racial/ethnic categories. For instance, among all Asian prospects (n=1,341) whose contact information was purchased in this SAT score range, approximately 12% were first-generation college students whose parents did not attend college, 16% were first-generation college students whose parents attended some college but did not complete their degree, and 72% were not first-generation college students. @fig-chicago-rq2-midSAT-combo shows the City of Chicago disproportionately contributes 35% of all Asian first-generation college students whose parents did not attend college. For first-generation college students whose parents attended some college but never completed degrees, City of Chicago (22%), South and Southwest Suburbs (20%), and Evanston and Skokie (20%) Geomarkets contributed larger shares of Asian prospects.

Similar Geomarket contribution patterns are evident among Hispanic prospects. The City of Chicago Geomarket stands out as the only Geomarket contributing substantially larger shares of Hispanic first-generation college students across those whose parents did not attend college (46% versus 38% overall) and those whose parents attended some college but did not complete their degrees (29% versus 26% overall). Although Black prospects who are not first-generation college students make up a relatively smaller portion of the overall pool, @fig-chicago-rq2-midSAT-combo suggests that such prospects are primarily contributed by the Evanston and Skokie, City of Chicago, and Western Surburbs Geomarkets.

Lastly, although Black prospects who are not first-generation college students make up a relatively smaller portion of the overall pool, @fig-chicago-rq2-midSAT-combo suggests that such prospects are primarily contributed by the Evanston and Skokie, City of Chicago, and Western Surburbs Geomarkets. Similarly, Geomarkets such as Chain of Lakes, North Shore, Evanston and Skokie, and Western Surburbs contribute a disproportionately larger share of White prospects who are not first-generation college students, whereas City of Chicago and South and Southwest Suburbs contribute larger shares of White first-generation college students.

@@

<!--# [MY READ IS THAT THE PATTERNS FOR WHITE ARE NOT THE SAME AS THE PATTERNS FOR ASIAN AND LATINX -- SO I WOULDN'T SAY THESE PATTERNS PERSIST ACROSS OTHER RACIAL/ETHNIC CATEGORIES. I WOULD RECOMMEND STARTING WITH ASIAN AND WITH HISPANIC. THAT IS THE LEDE; ALSO, THIS IS WHERE YOU COULD REFER TO THE SIMILAR GRAPH FOR PHILADELPHIA AND SAY IT FOLLOWS THE SAME PATTERN; I THINK THIS IS THE CLIMAX AND COOLEST INSIGHT OF THE RESULTS] KS=== I THINK ITS" WEIRD TO INCLUDE PHILLY GRAPH HERE WHEN ITS NOT CASE...WE CAN SPEAK TO THIS IN THE DISCUSSION INSTEAD... WE SEE SIMILAR PATTERNS ACROSS OTHER METROS TOO -->

Appendix figures @fig-chicago-rq2-lowSAT-combo and \@fig-chicago-rq2-highSAT-combo show the magnitude of these Geomarket contribution disparities become more pronounced at as SAT score thresholds increase.

The enduring segregated structure of the Chicago metropolitan area from 1980 to 2020 analyzed above plays a significant role in how Geomarkets shape the racial/ethnic and socioeconomic composition of prospect pools. Reflecting the larger metropolitan area dynamics, central Geomarkets that are predominantly comprised of low-income communities of color such as the City of Chicago and South and Southwest Suburbs consistently contribute disproportionately larger shares of Black, Hispanic, and first-generation college prospects, especially at lower SAT score thresholds. On the other hand, predominantly White and affluent Geomarkets in the suburban ring of the metropolitan area like the Chain of Lakes, North Shore, and Evanston and Skokie contribute larger proportions of White and Asian prospects, as well as non-first generation college students. However, the City of Chicago stands out in playing crucial role in fostering college access for first-generation Asian and Hispanic college students, contributing such prospects to the overall pool at significantly higher, disproportionate rates than any other Geomarket. Given the broader racial/ethnic distribution of the Chicago metropolitan area, student list orders that use Geomarket filters excluding the City of Chicago and including more affluent Geomarkets in targeting high-achieving students may still capture some Asian and Hispanic prospects. However, these students are far more unlikely to be first-generation students of color.

@@

From 1980 to 2020, the population of the overall Dallas-Fort Worth metropolitan area increased substantially from approximately 2.7 Million to 6.8 Million. @fig-dallas-maps shows all six Geomarkets experienced population growth over the three census periods. However, the most pronounced increases occured in the West of Dallas/Fort Worth Metroplex and Collin & Rockwall Counties Geomarkets, which grew from approximately 340,000 to 1.4 Million and 170,000 to 1.2 Million, respectively. These population trends, as shown in @fig-dallas-maps, suggest population growth has been partly driven by urban sprawl from the metropolitan centers - City of Forth Worth and City of Dallas - into the surrounding suburban and rural areas of the metropolitan areas. This outward expansion may also account for the comparatively larger geographic size of the West of Dallas/Fort Worth Metroplex Geomarket from 1980 to 2000.

<!--#  [WRITING IS QUITE GOOD!I THINK THERE IS A MODEST LEAD THAT THE NEXT COUPLE PARAGRAPHS DON'T STRESS ENOUGH , WHICH IS THAT TX19/CITY OF DALLAS IS THE MOST LEAST WHITE/ASIAN AND THE POOREST]--> @fig-dallas-rq1-race presents the racial and ethnic composition of each Geomarket within the Dallas-Fort Worth metropolitan area across the three census periods. In 1980, all Geomarkets were predominantly White, with White populations ranging from 62% in City of Dallas to nearly 92% in West of Dallas/Fort Worth Metroplex. By 2000, however, Geomarkets located in the central part of the metropolitan area exhibited a more balanced racial/ethnic distribution, including Dallas County (58% White, 16% Black, 19% Hispanic; 6% Asian), Irving, Arlington, and Grand Prairie (54% White, 12% Black, 26% Hispanic; 16% Asian), and City of Fort Worth (57% White, 16% Black, 22% Hispanic; 3% Asian). In contrast, Collin & Rockwall Counties and West of Dallas/Fort Worth Metroplex Geomarkets remained predominantly White in 2000 at 78% and 80%, respectively.

By 2000, the City of Dallas was no longer a predominantly White Geomarket, with a population that was approximately 24% Black, 30% Hispanic, and 3% Asian. Over the next two decades, the Hispanic population grew steadily, contributing to greater diversity. By 2020, the City of Dallas Geomarket had the lowest proportions of White and Asian residents across the Dallas-Fort Worth Metropolitan area, with approximately 21% Black, 38% Hispanic, and 2% Asian residents.

@@

>>>>

-->
```
# Results

Results are presented below for Chicago, Dallas, and Los Angeles. The [Online Appendix](https://ozanj.github.io/cb_geo/#/results) provides a more convenient means of viewing results. The online appendix also contains interactive maps, additional figures and tables not contained in the main text, and results for additional metropolitan areas (e.g., Bay Area, Philadelphia). The table of contents for navigating between metro areas can be accessed by clicking on the three horizontal lines on the bottom left of the screen.

@@

@fig-chicago-maps presents the seven Geomarkets in the Chicago metropolitan area by total population across the three census periods. The maps in the figure show how Chicago Geomarkets reflect the classic urban geographical structure of an inner-city core surrounded by a ring of suburbs as described by @RN4982[p. 12]. Based on the color gradient scale representing total population density, @fig-chicago-maps indicates northern suburb Geomarkets in the metropolitan area - Chain of Lakes, Northwest Suburbs, Northshore, Evanston and Skokie - experienced relatively little population growth from 1980 through 2020. In contrast, the most significant population increases occured in the Western Suburbs (1.2 Million to 1.6 Million) and the South and Southwest Suburbs (less than 1 Million to 1.3 Million) Geomarkets. The City of Chicago Geomarket, forming the metropolitan area's central core, was the only locality to experience a population decline, from 3.1 Million in 1980 to 2.8 Million in 2020, despite remaining the most populous Geomarket.

These modest population shifts belie substantial change in the racial/ethnic composition within and across Geomarkets over time. @fig-chicago-rq1-race illustrates the racial and ethnic composition of each Geomarket across the three census periods. In 1980, nearly all Geomarkets were predominantly White, with proportions ranging from 83% in the South and Southwest Suburbs to approximately 93% in North Shore. In contrast, the City of Chicago had a substantially more diverse population, with White residents comprising only 46% of its total population. One data caveat is that the following race/ethnicity categories were not available in 1980: Asian, non-Hispanic; Two+ races, non-Hispanic; NHPI, non-Hispanic; AIAN non-Hispanic. By 2020, several Geomarkets that were previously predominantly White shifted to a more heterogeneous racial/ethnic distribution, including Chain of Lakes, Northwest Suburbs, Evanston and Skokie, Western Suburbs, and South and Southwest Suburbs. For many Geomarkets, this balance in racial/ethnic composition was the result of large increases in the Latinx population from 1980 to 2020, which ranged from a 7 percentage point increase for the Evanston and Skokie Geomarket to a 19 percentage point increase for Western Suburbs. Some Geomarkets also experienced a substantial increase in Black populations (South and Southwest Suburbs) and in Asian populations (Northwest Suburbs and Evanston and Skokie). @fig-chicago-rq1-race also indicates that, although the North Shore and City of Chicago Geomarkets underwent some changes in racial and ethnic composition, their overall demographic patterns remained relatively the same across the three census periods. <!-- For instance, North Shore experienced some increases in Hispanic (2% in 1980 to 8% in 2020), Asian Pacific Islander (7% in 2000 to 10% in 2020), and Black (2% in 1980 to 3% in 2020) populations yet remained a predominantly White Geomarket (74% in 2020). City of Chicago also experienced relatively modest declines in its Black population (37% in 1980 to 33% in 2020), alongside increases in its Hispanic (15% in 1980 to 26% in 2020) and Asian Pacific Islander populations (4% in 2000 to 6% in 2020), further substantiating the Geomarket as the most racially/ethnic diverse in the metropolitan area. -->

The interactive maps in the [Online Appendix](https://ozanj.github.io/cb_geo/#/research-question-1-15) can show these shifts at the Census Tract-level. For example, in 1980 Black people were concentrated in southern tracks of the City of Chicago Geomarket. As described by scholarship on gentrification in Chicago [@ewing2024_nofuture; @snidal2022_exodus]. However, in 2000 and 2020 Black populations shift towards northeast tracks of the South and Southwest Suburbs Geomarket

We explore socioeconomic demographics -- median income, percent living below poverty, and educational attainment -- within and across Geomarkets over time in @fig-chicago-rq1-ses. The primary takeaways are the the City of Chicago -- and to a somewhat lesser degree th South and Southwest Suburbs - stands out as having lower median income, higher poverty rates, and lower BA attainment than other Geomarkets. By contrast, the North Shore stands outas being particularly affluent and highly educated. Overall, median income both within and across Geomarkets in the Chicago metropolitan area remained relatively stable across the three census periods. For instance, in 1980, median household income across the Chicago metropolitan area ranged from a high of \$158,000 in the North Shore Geomarket to a low of \$66,000 in the City of Chicago Geomarket. However, median incomes in all other Geomarkets were closer to the upper thresholds of affluence, such as \$120,000 in the Northwest Suburbs, \$119,000 in Evanston and Skokie, \$108,000 in Chain of Lakes, \$105,000 in the Western Suburbs, and \$99,000 in the South and Southwest Suburbs. While overall income patterns across Geomarkets remained consistent through 2020, a median income increase for the City of Chicago Geomarket (\$79,000) and decrease in the South and Southwest Suburbs Geomarket (\$93,000) narrowed the income gap between the lowest and second-lowest Geomarkets to just \$14,000. @fig-chicago-rq1-ses also shows that from 1980 to 2020, the percentage of residents living in poverty declined in the City of Chicago Geomarket while it increased in all other Geomarkets, resulting in a narrowing of poverty gaps across the Chicago metropolitan area. Similar patterns are evident in educational attainment across Geomarkets. In 1980, the City of Chicago had the lowest percentage of residents with a bachelor's degree or higher (13%). However, by 2020, the Geomarket experienced the largest percentage point increase in college degree attainment, reaching 38%.

Overall, demographic shifts transformed the Chicago-land Geomarket landscape from largely a binary structure into a more tiered dynamic that retains by substantial racial/ethnic and socioeconomic inequality between Geomarkets. The original Geomarket boundaries delineated significant racial and income segregation in the 1980s. This included Black, Hispanic, and Asian residents primarily in the City of Chicago Geomarket and White residents in Geomarkets forming the suburban ring of the metropolitan area. Income, poverty, and educational attainment also created a static socioeconomic divide in 1980, with southern Geormarkets (City of Chicago and South and Southwest Suburbs) characterized by lower affluence and educational attainment compared to the higher socioeconomic standing of the northwestern suburbs. However, declining White populations, Hispanic and Asian population growth, and Black out-migration, as well as narrowing of education and poverty gaps within some Geomarkets, have contributed to a more stratified Geomarket dynamic in the metropolitan area with significant disparities remaining between Geomarkets. The City of Chicago Geomarket remains the critical geographical hub across all Geomarkets for racial/ethnic diversity; however, other Geomarkets now also include substantive numbers of residents across racial/ethnic groups (such as South and Southwest Geomarket for Black residents, Chain of Lakes for Hispanic residents, and Northwest Suburbs for Asian residents). While some stark socioeconomic gaps within Geomarkets have narrowed since 1980, significant geographical income segregation remains between Geomarkets despite a more clearly defined middle-to-high socioeconomic Geomarket group (Chain of Lakes, Northwest Suburbs, Western Suburbs).

@@

To examine how the socioeconomic and racial composition of included versus excluded prospects varies when student list purchases filter on particular Geomarkets, we analyze six orders -- placed by a research university -- that target prospects across the entire state of Illinois. The first two orders filter on the same SAT thresholds (ranging from 1020 to 1150), the same GPA thresholds (ranging from B- to A+), and filter on the high school graduating classes of 2020 and 2021 graduating classes, respectively. The remaining four orders used the same GPA and high school class filters but differed in SAT thresholds. Two orders targeted students with relatively middle-range scores (1160 to 1300) and two orders targeted those with relatively high scores (1310 to 1600). Because all six orders included an Illinois state filter, we can utilize prospects from the Chicago metropolitan area within the resulting student lists to examine Geomarket contributions by simulating the application of a Geomarket filter across low-, middle-, and high-range SAT threshold orders.

@@

As PSAT score ranges increase, The City of Chicago Geomarket becomes the only Geomarket with substantial shares of Black, and Hispanic prospects. Moving from the top to the bottom panels, all Geomarkets experience growth in the shares of White and Asian prospects, while Black and Hispanic prospect shares decline. For the highest SAT score range, the City of Chicago Geomarket pool is 18% Asian, 7% Black, 21% Hispanic, 5% Multiracial, making it the only Geomarket that is not predominantly White (49%). Prospect pools from all other Geomarkets in this high PSAT order are predominantly White and Asian, with Black and Hispanic shares ranging from 0.4% (North Shore) to 5% (South and Southwest Suburbs) of prospects and 6% (Northwest Suburbs) to 10% (South and Southwest Suburbs) of prospects, respectively. <!--# [LEDE -- BUT IN BOTH THE MEDIUM AND HIGH THRESHOULD, CHICAGO IS THE ONLY GEOMARKET WHERE A SUBSTANTIAL SHARE OF PROSPECTS ARE NOT WHITE OR ASIAN] -->

We also explore contributions across Geomarkets to overall prospect pools by race/ethnicity. @fig-chicago-rq2-midSAT-race shows the contributions of each Geomarket to the pool of prospects from the two orders targeting middle range SAT scores (1160-1300) by race/ethnicity. The top panel presents the full distribution of the 13,643 total prospects (whose race/ethnicity is known) included in these orders by Geomarkets: 11% of all purchased prospect profiles reside in Chain of Lakes, 13% in Northwest Suburbs, 5% in North Shore, 5% in Evanston and Skokie, 17% in City of Chicago, 28% in Western Suburbs, and 20% in South and Southwest Suburbs. These overall distributions are compared to those in the lower panels of @fig-chicago-rq2-midSAT-race, which illustrate the proportional representation of each race/ethnicity group across Geomarkets.

@@

@fig-chicago-rq2-highSAT-combo shows the magnitude of Geomarket contribution disparities increases at the highest SAT score range (1310-1600), partly driven by the relatively fewer number of first-generation college students included within these orders. Again, the City of Chicago Geomarket disproportionately contributes a disproportionate share of first-generation Asian college students (43% across both parental education categories) and first-generation Hispanic college students (52% across both parental education categories) in comparison to the overall distribution of first-generation college students within each racial/ethnic category (13% for Asian and 40% for Hispanic, respectively). Conversely, nearly all other Geomarkets contribute a disproportionately larger share of non-first generation college students across all racial/ethnic categories. -->

@fig-chicago-rq2-midSAT-race shows which Geomarkets send the bulk of students for a given race/ethnicity. The City of Chicago accounts for 17% of all prospects but 48% of Black students. Similarly, South \& Southwest Suburbs accounts for 20% of all prospects but 30% of all Black prospects. The upshot is that if the student list purchases had excluded these two Geomarkets, 78% of the nearly 1,000 Black students currently included in these purchases would be excluded. On the other hand, all other Geomarkets contributed substantially smaller shares of Black prospects. For example, the Northwest Suburbs accounts for about 13% of all prospects but only about 3% of Black prospects. The City of Chicago Geomarket also contributed larger shares of the nearly 3,000 Hispanic students (39% versus 17% for all prospects), while all other Geomarkets contributed comparable or relatively smaller proportions of Hispanic prospects than their overall representation among all included prospects.

In contrast, these representational patterns by Geomarket are reversed for White and Asian students. @fig-chicago-rq2-midSAT-race shows nearly all Geomarkets contributed to the pool of White prospects at rates proportional to their overall contributions to all included prospects (except City of Chicago). Similar patterns are observed for Asian students, with proportional contributions by Geomarket largely consistent with their overall distribution, except for Northwest Suburbs (19% versus 13% for all prospects) and Evanston and Skokie (10% versus 5% for all prospects) contributing disproportionately larger shares and the South and Southwest Suburbs (8% versus 10% for all prospects) contributing smaller shares relative to their overall contribution to the larger prospect pool.

@@

<!--KARINA -- I AM CUTTING THIS. # Relative to @fig-chicago-rq2-midSAT-ses, appendix figures \@fig-chicago-rq2-midSAT-ses and @fig-chicago-rq2-highSAT-ses demonstrate Geomarket contribution patterns by first-generation college student status are less pronounced in low-SAT range orders (1020-1150) and more pronounced in high-range SAT orders (1310-1600), respectively. -->

Finally, we examine Geomarket contributions to prospect pools by both first-generation college student status and race/ethnicity in the Chicago metropolitan area. @fig-chicago-rq2-midSAT-combo presents the Geomarket distribution of the 12,981 total prospects (whose first-generation status and race/ethnicity is known) included in middle-range SAT (1160-1300) orders. @fig-chicago-rq2-midSAT-combo shows the City of Chicago Geomarket consistently contributes a disproportionately larger share of first-generation college students whose parents did not attend college across all racial/ethnic categories. For instance, among all Asian prospects (n=1,341) whose contact information was purchased in this SAT score range, approximately 12% were first-generation college students whose parents did not attend college, 16% were first-generation college students whose parents attended some college but did not complete their degree, and 72% had parents with a BA. By contrast, for 237 Asian students from the City of Chicago Geomarket, 35% had parents who did not attend colllege and only 43% had parents with a BA.

Similar Geomarket contribution patterns are evident among Hispanic prospects. The City of Chicago Geomarket stands out as the only Geomarket contributing substantially larger shares of Hispanic first-generation college students across those whose parents did not attend college (46% versus 38% overall) and those whose parents attended some college but did not complete their degrees (29% versus 26% overall). By contrast, Hispanic students from the more affluent North Shore Geomarket and Evanston \& Skokie Geomarket were more likely to have parents with a BA.

Although Black prospects who are not first-generation college students make up a relatively smaller portion of the overall pool, @fig-chicago-rq2-midSAT-combo suggests that such prospects are primarily contributed by the Evanston and Skokie, City of Chicago, and Western Surburbs Geomarkets. Similarly, Geomarkets such as Chain of Lakes, North Shore, Evanston and Skokie, and Western Surburbs contribute a disproportionately larger share of White prospects who are not first-generation college students, whereas City of Chicago and South and Southwest Suburbs contribute larger shares of White first-generation college students.

@@

<!--# [MY READ IS THAT THE PATTERNS FOR WHITE ARE NOT THE SAME AS THE PATTERNS FOR ASIAN AND LATINX -- SO I WOULDN'T SAY THESE PATTERNS PERSIST ACROSS OTHER RACIAL/ETHNIC CATEGORIES. I WOULD RECOMMEND STARTING WITH ASIAN AND WITH HISPANIC. THAT IS THE LEDE; ALSO, THIS IS WHERE YOU COULD REFER TO THE SIMILAR GRAPH FOR PHILADELPHIA AND SAY IT FOLLOWS THE SAME PATTERN; I THINK THIS IS THE CLIMAX AND COOLEST INSIGHT OF THE RESULTS] KS=== I THINK ITS" WEIRD TO INCLUDE PHILLY GRAPH HERE WHEN ITS NOT CASE...WE CAN SPEAK TO THIS IN THE DISCUSSION INSTEAD... WE SEE SIMILAR PATTERNS ACROSS OTHER METROS TOO Appendix figures @fig-chicago-rq2-lowSAT-combo and \@fig-chicago-rq2-highSAT-combo show the magnitude of these Geomarket contribution disparities become more pronounced at as SAT score thresholds increase. -->

In summary, the enduring segregated structure of the Chicago metropolitan area from 1980 to 2020 plays a significant role in how Geomarkets shape the racial/ethnic and socioeconomic composition of prospect pools. Reflecting the larger metropolitan area dynamics, Geomarkets that are predominantly comprised of low-income communities of color such as the City of Chicago and South and Southwest Suburbs consistently contribute disproportionately larger shares of Black, Hispanic, and first-generation college prospects, especially at lower SAT score thresholds. On the other hand, predominantly White and affluent Geomarkets in the suburban ring of the metropolitan area like the Chain of Lakes, North Shore, and Evanston and Skokie contribute larger proportions of White and Asian prospects, as well as non-first generation college students. However, the City of Chicago stands out in playing crucial role in fostering college access for first-generation Asian and Hispanic college students, contributing such prospects to the overall pool at significantly higher, disproportionate rates than any other Geomarket. Given the broader racial/ethnic distribution of the Chicago metropolitan area, student list orders that use Geomarket filters excluding the City of Chicago and including more affluent Geomarkets in targeting high-achieving students may still capture some Asian and Hispanic prospects. However, these students are far more unlikely to be first-generation students of color.

@@

From 1980 to 2020, the population of the overall Dallas-Fort Worth metropolitan area increased dramatically from approximately 2.7 Million to 6.8 Million. @fig-dallas-maps shows all six Geomarkets experienced population growth over the three census periods. However, the most pronounced increases occured in the West of Dallas/Fort Worth Metroplex and Collin & Rockwall Counties Geomarkets, which grew from approximately 340,000 to 1.4 Million and 170,000 to 1.2 Million, respectively. These population trends, as shown in @fig-dallas-maps, suggest population growth has been partly driven by urban sprawl from the metropolitan centers -- City of Forth Worth and City of Dallas -- into the surrounding suburban and rural areas of the metropolitan areas. This outward expansion may also account for the comparatively larger geographic size of the West of Dallas/Fort Worth Metroplex Geomarket from 1980 to 2000.

<!--#  [WRITING IS QUITE GOOD!I THINK THERE IS A MODEST LEAD THAT THE NEXT COUPLE PARAGRAPHS DON'T STRESS ENOUGH , WHICH IS THAT TX19/CITY OF DALLAS IS THE MOST LEAST WHITE/ASIAN AND THE POOREST]--> @fig-dallas-rq1-race presents the racial and ethnic composition of each Geomarket within the Dallas-Fort Worth metropolitan area across the three census periods. In 1980, all Geomarkets were predominantly White, with White populations ranging from 62% in City of Dallas to nearly 92% in West of Dallas/Fort Worth Metroplex. By 2000, however, Geomarkets located in the central part of the metropolitan area exhibited a more balanced racial/ethnic distribution, including Dallas County, Irving, Arlington, and Grand Prairie, and City of Fort Wort. In contrast, Collin & Rockwall Counties and West of Dallas/Fort Worth Metroplex Geomarkets remained predominantly White in 2000 at 78% and 80%, respectively. By 2000, the City of Dallas was no longer a predominantly White Geomarket, with a population that was approximately 24% Black, 30% Hispanic, and 3% Asian. Over the next two decades, the Hispanic population grew steadily, contributing to greater diversity. By 2020, the City of Dallas Geomarket had the lowest proportions of White and Asian residents across the Dallas-Fort Worth Metropolitan area, with approximately 21% Black, 38% Hispanic, and 2% Asian residents.

@@

\begingroup
\small
